---
title: "STA380_James_Problems"
author: "Grayson Merritt"
date: "2024-07-31"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1
We are looking for P(Yes|TC)
P(Yes|RC) = .5
P(No|RC) = .5
P(Yes) = .65
P(No) = .35
P(RC) = .3

The total law of probability states that P(A) = the summation of P(A|B) * P(B)
The P(Yes) comes from only two conditional probabilities: P(Yes|RC) and P(Yes|TC)
So P(Yes)=P(Yes∣RC)P(RC)+P(Yes∣TC)P(TC)
Using some algebra I can arrange this to (P(Yes) - P(Yes|RC)P(RC)) / P(TC) = P(Yes∣TC)
Thus, P(Yes|TC) is **71.43%**

### Part B
Sensitivity = P(P|D) = .993
Specificity = P(N|ND) = .9999
Disease = P(D) = .000025
The question we are solving is: What is the P(D|P)? 
Bayes Theorem is P(D|P) = (P(P|D)*P(D)) / P(P)
We have P(P|D) and P(D), so we need to find P(P). This will require using the rule of total probability 
So P(P) = P(P|D) * P(D) + P(P|ND) * P(ND)
So we need P(ND) and P(P|ND)
No disease = P(ND) = 1 - P(D) = .999975
P(P|ND) = 1 - P(N|ND) = .0001  (This is the False Positive case)
P(P) = .00012
After calculating all of my needed info and applying Bayes Theorem, I get that the Probability of a person having the disease given a positive test is *19.88%*

```{r Problem 1}
p_yes_rc = .5
p_no_rc = .5
p_yes = .65
p_no = .35
p_rc = .3
p_tc = .7
p_yes_tc = (p_yes-p_yes_rc*p_rc) / p_tc
print(p_yes_tc)

# Part B
sensitivity = .993
specificity = .9999
disease = .000025
no_disease = 1- disease
no_disease
false_postive = 1 - specificity
false_postive
positive = sensitivity * disease + false_postive * no_disease
positive
disease_given_positive = (sensitivity * disease)/ positive
print(disease_given_positive)

```

# Question 2

### Part A
This table shows the top ten most popular songs since 1958 based on how long they were
on the billboard 100. Most of these songs were produced in the last 21 years. I find it
interesting that there are no repeats of performers on this top ten list.

```{r Problem 2, echo=FALSE}
library(tidyverse)
library(readr)
billboard = read_csv("billboard.csv")
#question 2
#a 
billboard_a = billboard %>% group_by(performer,song)
billboard_a %>% summarize(count = n()) %>% arrange(desc(count))
```
### Part B
This plot shows the total number of unique songs that charted the Billboard 100 per year. I guess my
parents were correct when they said music was better in the 80’s! The number of unique songs that chart
peaks at around 1967 and then rapidly declines until around 2002, where more unique songs started to chart.
This could potentially be due to the rise if iTunes. 
There was a decline around 2011 followed by rapid unique song growth.


```{r Problem 2 Part B}
#b
billboard_cutoff = billboard %>% filter(year != 1958 & year != 2021)
table_with_counts = billboard_cutoff %>% group_by(performer,song,year) %>% 
  summarize(total_count = n())
unique_song_count = table_with_counts %>% group_by(year) %>% 
  summarize(unique_songs = n())
ggplot(unique_song_count) + geom_line(aes(x=year,y=unique_songs))
```
### Part C
This plot shows artists who have had 30 songs chart for at least ten weeks. Elton John has the highest
number of songs with 52 songs. I find it interesting that there are a good amount of country artists filled. I
would have thought this list would have been mainly filled with pop and rock artists
```{r Problem 2 Part C}
#C
billboard_ten_week = billboard %>% group_by(performer,song) %>% 
  summarize(count = n()) %>%
  filter(count >=10)
billboard_19_artists = billboard_ten_week %>% group_by(performer) %>% 
  summarize(song_count =n()) %>% filter(song_count >=30)
ggplot(billboard_19_artists) + geom_col(aes(x=performer, y=song_count)) + 
  coord_flip()
```
# Problem 3
In order to agree with the stats guru's conclusion that building a green building makes sense,
we first have to do our own analysis of the data and see what findings we can come up with. 
We first took a look at the median rent of green buildings vs non green buildings and confirmed that 
on average green buildings earn about $2.6 more per sq ft than non green buildings. 
Our general strategy is  to see what  factors could affect rent and see if these factors are 
over or under represented in the green buildings. 
We then looked to see if green buildings tended to be more of Class A. We found that around 80%
of green rated buildings are class A, versus only 36% of non green buildings. This may be a 
potential confounder, as class A buildings will command more rent. 
We plotted Rent vs age and found that an older building commands less rent. However,
the non green buildings are almost 50 years old on average vs 24 years for green buildings!
This could certainly be a confounder. 
Rent is LOWER for renovated buildings. This is not what we expected. Renovated buildings
command less rent by $3.79! Most green buildings (around 79%), however, 
are NOT renovated! This means that they are drawing higher prices due to not being renovated. 

So far we have found 3 plausible explanations for why green buildings demand higher rent. The class A
variable is likely one of the biggest confounders, so we want to see the median rent within non class A buildings
faceted by green rating. We found that after adjusting for class A buildings the premium is only $2.12 now.

We think this trend will continue for most of these confounders. We think the best way to adjust for these 
confounders is to use a technique called matching. Matching relies on a simple principle: compare like with like.
In this example, that means if we have a 25-year-old, Class A building that is renovated with a green rating, 
we try to find another 25-year old, Class A renovated building without a green rating to compare it to.
Matching  constructs a balanced data set from an unbalanced one. This matched data can them be compared by their rents
to see if green buildings truly cause a higher premium. 

We think the stats guru did not take into account any confounders in his model. While green builfsing may
deman higher rent, we think there any more variables at play here. 


```{r}
library(mosaic)
greenbuildings = read_csv("greenbuildings.csv")
# also remove the occupancy rate of less than 10%
greenbuildings = greenbuildings %>% filter('leasing_rate' >= 10)

median(Rent ~ green_rating, data=greenbuildings)

ggplot(greenbuildings) + 
  geom_boxplot(aes(x=factor(green_rating), y=Rent)) + 
  coord_flip()

# Look at which buildings are "more desirable" (Class A)
xtabs(~ class_a + green_rating, data=greenbuildings) %>%
  prop.table(margin=2)

# Look at how age affects rent
mean(age ~ green_rating, data=greenbuildings)

ggplot(greenbuildings, aes(x = age, y = Rent)) +
  geom_point(aes(color = as.factor(green_rating)), alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Rent vs Age of Buildings",
       x = "Age of Building (years)",
       y = "Rent ($ per square foot per year)",
       color = "Green Rating") +
  theme_minimal()

# look to see if renovated affects rent
mean(Rent ~ renovated, data = greenbuildings)
xtabs(~ renovated + green_rating, data=greenbuildings) %>%
  prop.table(margin=2)



# Look at non class A buildings 
median_rent_non_class_a <- greenbuildings %>%
  filter(class_a == 0) %>%  # Exclude Class A buildings
  group_by(green_rating) %>%
  summarise(median_rent = median(Rent))

print(median_rent_non_class_a)

ggplot(median_rent_non_class_a, aes(x = factor(green_rating), y = median_rent, fill = factor(green_rating))) +
  geom_bar(stat = "identity") +
  labs(title = "Median Rent for Non-Class A Buildings by Green Rating",
       x = "Green Rating (0 = Non-Green, 1 = Green)",
       y = "Median Rent ($ per square foot per year)",
       fill = "Green Rating") +
  scale_fill_manual(values = c("0" = "grey", "1" = "green")) +
  theme_minimal()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
