---
title: "STA380_James_Problems"
author: "Grayson Merritt"
date: "2024-07-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1
We are looking for P(Yes|TC)
P(Yes|RC) = .5
P(No|RC) = .5
P(Yes) = .65
P(No) = .35
P(RC) = .3

The total law of probability states that P(A) = the summation of P(A|B) * P(B)
The P(Yes) comes from only two conditional probabilities: P(Yes|RC) and P(Yes|TC)
So P(Yes)=P(Yes∣RC)P(RC)+P(Yes∣TC)P(TC)
Using some algebra I can arrange this to (P(Yes) - P(Yes|RC)P(RC)) / P(TC) = P(Yes∣TC)
Thus, P(Yes|TC) is **71.43%**

# Part B
Sensitivity = P(P|D) = .993
Specificity = P(N|ND) = .9999
Disease = P(D) = .000025
The question we are solving is: What is the P(D|P)? 
Bayes Theorem is P(D|P) = (P(P|D)*P(D)) / P(P)
We have P(P|D) and P(D), so we need to find P(P). This will require using the rule of total probability 
So P(P) = P(P|D) * P(D) + P(P|ND) * P(ND)
So we need P(ND) and P(P|ND)
No disease = P(ND) = 1 - P(D) = .999975
P(P|ND) = 1 - P(N|ND) = .0001  (This is the False Positive case)
P(P) = .00012
After calculating all of my needed info and applying Bayes Theorem, I get that the Probablity of a person having the disease given a positive test is *.1989*
```{r Problem 1}
p_yes_rc = .5
p_no_rc = .5
p_yes = .65
p_no = .35
p_rc = .3
p_tc = .7
p_yes_tc = (p_yes-p_yes_rc*p_rc) / p_tc
p_yes_tc

# Part B
sensitivity = .993
specificity = .9999
disease = .000025
no_disease = 1- disease
false_postive = 1 - specificity
positive = sensitivity * disease + false_postive * no_disease
positive
disease_given_positive = (sensitivity * disease)/ positive
disease_given_positive

```

## Including Plots

You can also embed plots, for example:

```{r Problem 2, echo=FALSE}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
